---
title: "Riddler Gematria"
author: "Alex Rossell Hayes"
date: "January 10, 2020"
output: github_document
---

# The problem

This Riddler sets up an English version of Gematria, in which letters are scored
based on their place in the alphabet.
A is 1, B is 2, etc.
Our task is to find the largest number which, when expressed in words, has a
gematria score higher than its numerical value.
'One', with a score of 34, fits this criteria, while 'one thousand', with a
score of 136, does not.

# Creating a function

In order to solve this, we need to create a function that finds the English
Gematria score for a given number.

The first step in achieving this uses the package `english`, which allows us to
convert an integer to its representation in English words.

```{r}
library(english)

as.english(52)
```

Because the Gematria system doesn't give any value to punctuation characters,
let's get rid of them using the `stringr` package.
The regular expression "\\W" finds all 'non-word' characters, meaning anything
that is not a letter, digit, or underscore.

```{r}
library(stringr)

as.english(52) %>%
  str_remove("\\W")
```

Because scores are calculated at the individual character level, let's split the
word from a single string to a vector of characters.
`str_split()` is designed to split one string into multiple strings at a
specified break point.
For example, a month-day-year could be split at the dashes:

```{r}
str_split("01-10-2020", "-")
```

However, by using an empty string ("") as our split pattern, we can divide the
string into individual characters.
This puts the string into a list, so let's also call `unlist()` to make it a
single vector.

```{r}
as.english(52) %>%
  str_remove("\\W") %>%
  str_split("") %>%
  unlist()
```

We need to find each letter's place in the alphabet.
To do that, we can compare the letters to R's built-in `letters` variable, which
contains each letter in alphabetical order.

```{r}
letters
```

To find a letter's place, we can use the `which()` function to find which
character in letters equals our desired character.

```{r}
which(letters == "g")
```

To find the values for each character in our vector, we can use the `furrr`
package, which allows us to apply a function to each element in a set.
`furrr` works just like the better-known `purrr` package, but uses parallel
computing processes to speed up our analysis

```{r}
library(furrr)
plan(multiprocess) # Tell furrr to use multiple processes to speed up analysis

as.english(52) %>%
  str_remove("\\W") %>%
  str_split("") %>%
  unlist() %>%
  future_map_int(~ which(letters == .))
```

Finally, we use `sum()` to add up the values, giving us the Gematria score for
the whole word.

```{r}
as.english(52) %>%
  str_remove("\\W") %>%
  str_split("") %>%
  unlist() %>%
  future_map_int(~ which(letters == .)) %>%
  sum()
```

Putting this all together, we can make a function that calculates the Gematria
score for an arbitrary number:

```{r}
gematria <- function(integer) {
  integer %>%
    as.english() %>%
    str_remove_all("\\W") %>%
    str_split("") %>%
    unlist() %>%
    future_map_int(~ which(letters == .)) %>%
    sum()
}

gematria(538)
```

# Narrowing the range

Now that we have a function, we can apply it to a list of numbers to get our
answer.
But the question asks for the largest number that fits our criteria.
With infinite numbers, we can't exhaustively test them all.
Let's try to set an upper limit for our analysis.

The most points that can be scored by a single digit is the 65 points scored by
7:

```{r}
future_map_int(1:9, gematria) %>% `names<-`(1:9)
```

Similarly, 70 does best in the tens place:

```{r}
future_map_int(1:9 * 10, gematria) %>% `names<-`(1:9 * 10)
```

It seems then that the maximum score for each number of digits will be scored by
a string of successive 7s.
Let's see how they score for one to four digits:

```{r}
future_map_int(c(7, 77, 777, 7777), gematria) %>% `names<-`(c(7, 77, 777, 7777))
```

7 and 77 have higher scores than their values, while 777 and 7777's scores are
too low.

This shows us that all three and four digit numbers do not have scores that are
more than three digits.
It also shows us that this function has diminishing returns: each 10-fold
increase only increases the score by less than 200.

Given this information, we can be confident that the answer must be somewhere
between 77 and 777.
This gives us a reasonable range in which to test our function.

# The analysis

Using `dplyr`, we can make a tibble with a row for each number between 777 and
77 (descending because we're interested in the highest number).

```{r}
library(dplyr)

(df <- tibble(number = 777:77))
```

For each number, we `mutate()` to add a column with its Gematria score.

```{r}
(df <- df %>% mutate(gematria_score = future_map_int(number, gematria)))
```

Finally, we `filter()` our results to only include those with a Gematria score
greater than the numerical value.

```{r}
(df <- df %>% filter(gematria_score > number))
```

We see that the highest number left in the tibble is 279, so this is our answer!

# The answer

**279** is the largest number with a Gematria score larger than its numerical
value.